---
title: "formula_algorithms"
author: "Haolin Zhong"
date: "2022/4/27"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


# The Posterior Distribution of the Parameters $\Theta$

- Assumption:

$$
Y_{i}(t+6)=\beta_{0, i}+\beta_{1, i} Y_{i}(t)+\beta_{2, i} \Delta_{i, 1}(t)+\beta_{3, i} \Delta_{i, 2}(t)+\beta_{4, i} \Delta_{i, 3}(t)+\epsilon_{i}(t)
$$

\vskip 10pt

$$
f(\boldsymbol{B} \mid \boldsymbol{\beta},\boldsymbol{\Sigma})= \prod_{i=1}^{n} \left\{ (2 \pi)^{-5 / 2}|\Sigma|^{-1 / 2} \exp \{-\frac{1}{2}(\boldsymbol{\beta}_i-\boldsymbol{\beta})^{\prime} \Sigma^{-1}(\boldsymbol{\beta}_i-\boldsymbol{\beta})\} \right\}
$$

\vskip 10pt


$$
P\left(\sigma^{2}\right) \propto \frac{1}{\sigma^{2}} ; \quad P(\boldsymbol{\beta}) \propto 1 ; \quad P\left(\Sigma^{-1}\right) \propto|\Sigma|^{-(d+1)} \exp \left(-\frac{1}{2} \Sigma^{-1}\right)
$$

\vskip 10pt

- Let's denote:

$$
\eta_{i}(t) = \beta_{0, i}+\beta_{1, i} Y_{i}(t)+\beta_{2, i} \Delta_{i, 1}(t)+\beta_{3, i} \Delta_{i, 2}(t)+\beta_{4, i} \Delta_{i, 3}(t)
$$

\vskip 10pt

- Distribution of $\epsilon_{i}(t)$:

$$
\epsilon_{i}(t) \sim N(0, \sigma^2)
$$

\vskip 10pt

- Distribution of $Y_i(t+6)$

$$
Y_{i}(t+6) \sim N(\eta_{i}(t) , \sigma^2)
$$

\vskip 10pt

- pdf for $Y_i(t+6)$:

$$
f(Y_{i}(t+6) \mid \boldsymbol \beta_i, \boldsymbol \beta, \sigma^2, \boldsymbol \Sigma) = -\frac{1}{\sqrt{2\pi\sigma^2} } \exp\{-\frac{1}{2\sigma^2} [Y_{i}(t+6) - \eta_{i}(t) ]^2\}
$$

\vskip 10pt

- pdf for $Y$: ($t_{i(n)}$ is the time of the last record for hurricane i)


$$
f(Y \mid \boldsymbol B, \boldsymbol \beta, \sigma^2, \boldsymbol   \Sigma) = \prod_{i=1}^n \prod_{t=0}^{t_{i(n)} - 6} -\frac{1}{\sqrt{2\pi} \sigma} \exp\{-\frac{1}{2\sigma^2} [Y_{i}(t+6) - \eta_{i}(t) ]^2\}
$$

\vskip 10pt

- posterior distribution of the parameters $\theta$:

$$
P(\Theta \mid Y) \propto f(Y | \Theta)P(\Theta) = f(Y \mid \boldsymbol B, \boldsymbol \beta, \sigma^2, \boldsymbol \Sigma) 
f(\boldsymbol B \mid \boldsymbol \beta, \boldsymbol \Sigma)P(\boldsymbol \beta) P(\sigma^2) P(\boldsymbol \Sigma^{-1})
$$

\vskip 10pt



- for efficient computation, we take logarithm:

$$
\begin{aligned}
\log P(\Theta \mid Y) & \propto \sum_{i=1}^{n} \sum_{t=0}^{t_{i(n)} - 6} \{ \log{(-\sqrt{2\pi}\sigma)} -\frac{1}{2\sigma^2} [Y_{i}(t+6) - \eta_{i}(t) ]^2\}
\\ & + \sum_{i=1}^{n} \{-\frac{1}{2}(\boldsymbol{\beta}_i-\boldsymbol{\beta})^{\prime} \Sigma^{-1}(\boldsymbol{\beta}_i-\boldsymbol{\beta})\}
\\ & - \log(\sigma^2) - (d+1)\log(|\boldsymbol \Sigma|) - \frac{1}{2} \text{trace} (\boldsymbol \Sigma^{-1})
\end{aligned}
$$




\vskip 10pt


# a MCMC algorithm to generate the posterior distribution of $\Theta$

## Data Preparation

```{r}
library(tidyverse)
library(lubridate)

correct.year = function(date) {
  date$year = date$year - 100
  return(date)
}

raw = read_csv("data/hurrican703.csv") %>% 
  janitor::clean_names() %>% 
  select(-nature, -season, -month) %>% 
  mutate(
    time = gsub("[()]", "", time),
    time = as.POSIXlt(parse_datetime(time, "%y-%m-%d %H:%M:%S")),
    time = as.POSIXct(correct.year(time))
  ) %>% 
  filter(hour(time) %in% c(0, 6, 12, 18),
         minute(time) == 0,
         second(time) == 0)

placeholder = data.frame(id = "placeholder", 
                        time = parse_datetime("00-01-01 00:00:00", "%y-%m-%d %H:%M:%S"), 
                        latitude = 0, 
                        longitude = 0,
                        wind_kt = 0)

dt = bind_cols(rbind(raw, placeholder), 
          rbind(placeholder, raw), 
          .name_repair = "unique")  

dt = dt[2:(nrow(dt)-1),]

dt = dt %>% 
  filter(id...1 == id...6,
         time...7 + 6*60*60 == time...2) %>% 
  mutate(
    d_lat = latitude...3 - latitude...8,
    d_log = longitude...4 - longitude...9,
    d_wkt = wind_kt...5 - wind_kt...10
  ) %>% 
  select(id = id...1, wkt_new = wind_kt...5, wkt_cur = wind_kt...10, d_lat, d_log, d_wkt)

hc = distinct(dt, id) %>% add_rownames("i")

dt = dt %>% left_join(hc)

head(dt)
```







